The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-25 conf_article/neon.tex
provided proofs in section 4

Our experiments reveal important insights about the role of matrix norms in optimization. On a synthetic least squares problem, we observe a striking discrepancy: while some algorithms converge slowly in terms of loss, they may converge quickly in terms of gradient norm, and vice versa. This suggests that existing theoretical guarantees may not fully explain practical algorithm performance.

@@

\section{Duals to Ky Fan norms instead of the spectral norm}

@@

\section{Conic combination of LMO-algorithms is an LMO-algorithm}

NSGD, Muon, and SignSGD work fair enough separately. Could we boost the quality if we combined them? To answer this question, we have to build a theoretical base for that.

\subsection{Duals to Convex Combination of Norm}
First, we need a well-known fact mentioned, for example, in \citet[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof of the fact.

\begin{lemma}\label{lemma:dual_to_conv_comb}
	Let $\norm{\cdot}_{(1)}$ and $\norm{\cdot}_{(2)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha,\beta \geq 0$. Define
	$$
		\norm{x} := \alpha \norm{x}_{(1)} + \beta \norm{x}_{(2)}.
	$$
	Then the dual unit ball of $\norm{\cdot}$ satisfies
	$$
		B_{\norm{\cdot}^\dagger}
		= \alpha B_{\norm{\cdot}_{(1)}^\dagger} + \beta B_{\norm{\cdot}_{(2)}^\dagger},
	$$
	where $+$ denotes the Minkowski sum and $B_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

\begin{proof}
	Write $f(x) = \alpha \norm{x}_{(1)}$ and $g(x) = \beta \norm{x}_{(2)}$, so
	$$
		\norm{x} = f(x)+g(x).
	$$
	Recall two standard facts:
	\begin{enumerate}
		\item For any norm $\norm{\cdot}$ and $\lambda>0$,
		      $$
			      (\lambda \norm{\cdot})^*(y) =
			      \sup_{x}\bigl( \langle y,x \rangle - \lambda \norm{x}\bigr)
			      = \delta_{\lambda B_{\norm{\cdot}^\dagger}}(y),
		      $$
		      i.e., the indicator function of the scaled dual ball.
		\item The Fenchel conjugate of a sum satisfies
		      $$
			      (f+g)^*(y) = \inf_{u+v=y} \bigl(f^*(u) + g^*(v)\bigr).
		      $$
	\end{enumerate}
	Applying these to $f$ and $g$, we have
	$$
		f^*(u) = \delta_{\alpha B_{\norm{\cdot}_{(1)}^\dagger}}(u),
		\quad
		g^*(v) = \delta_{\beta B_{\norm{\cdot}_{(2)}^\dagger}}(v).
	$$
	Thus
	$$
		\norm{\cdot}^*(y)
		= (f+g)^*(y)
		= \inf_{u+v=y}
		\bigl(
		\delta_{\alpha B_{\norm{\cdot}_{(1)}^\dagger}}(u)
		+
		\delta_{\beta B_{\norm{\cdot}_{(2)}^\dagger}}(v)
		\bigr)
		=
		\delta_{\alpha B_{\norm{\cdot}_{(1)}^\dagger}+\beta B_{\norm{\cdot}_{(2)}^\dagger}}(y).
	$$
	But by definition, the conjugate of a norm is exactly the indicator of its dual unit ball:
	$$
		\norm{\cdot}^*(y) = \delta_{B_{\norm{\cdot}^\dagger}}(y).
	$$
	Therefore,
	$$
		B_{\norm{\cdot}^\dagger} = \alpha B_{\norm{\cdot}_{(1)}^\dagger} + \beta B_{\norm{\cdot}_{(2)}^\dagger}.
	$$
\end{proof}

Consequently,
\begin{equation}\label{eq:dual_to_conv_comb}
	\norm{y}^\dagger
	= \inf\bigl\{ t \ge 0 : y \in t\bigl(\alpha B_{\norm{\cdot}_{(1)}^\dagger} + \beta B_{\norm{\cdot}_{(2)}^\dagger}\bigr) \bigr\}
	= \inf_{z,t}\bigl\{t \ge 0: z \in t \alpha B_{\norm{\cdot}_{(1)}^\dagger}, y - z \in t \beta B_{\norm{\cdot}_{(1)}^\dagger}\bigr\}
\end{equation}

Now we can work with different norms.

\subsection{General case}
We have norm \#1 with update \#1 and norm \#2 with update \#2. We prove the lemma that says that the update of their conic combination is a conic combination of updates \#1 and \#2.

Then prove a lemma for any set of norms \#1, \dots, \#n (by induction)

\subsection{Frobeniusize them: F-Muon and F-Neon}
Let us consider a mixture of Fanions with another excellent LMO-algorithm, NSGD.

We define $\normfstar{\cdot}$ as a convex combination of $\normn{\cdot}$ and $\normf{\cdot}$:

@@

\section{Computing the updates}\label{sec:matrix-side}
To compute the algorithms' updates, we use thick-restarted Lanczos method for singular value problem (TRLan) on $\mM^{k\top}\mM^k$ or $\mM^k \mM^{k\top}$ matrices (the one with less size is picked), implemented in CuPy lSibrary \citep{cupy_svds_ref} and described in \citet{simonz1998thick}.

>>>>

Our experiments reveal important insights about the role of matrix norms in optimization. \textcolor{red}{On a synthetic least squares problem, we observe a striking discrepancy: while some algorithms converge slowly in terms of loss, they may converge quickly in terms of gradient norm, and vice versa}. This suggests that existing theoretical guarantees may not fully explain practical algorithm performance.

@@

\section{Duals to Ky Fan Norms Instead of the Spectral Norm}

@@

\section{Conic Combination of LMO-algorithms is an LMO-algorithm}

Simply applying norm dual to the Ky Fank $k$-norm however, is not enough to provide family of algorithms diverse enough for our cause. Next we consider linear combinations of algorithms from LMO framework, which happen to be LMO-algorithms too, as we show in succeeding paragraphs.

\subsection{General Case}
First, we need a well-known fact mentioned, for example, in \citet[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof here.

\begin{lemma}\label{lemma:dual_to_conv_comb}
	Let $\norm{\cdot}_{(1)},\dots,\norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1,\dots,\alpha_n$ be non-negative reals. Define
	$$
		\norm{x} := \sum_{i = 1}^n \alpha_i \norm{x}_{(i)}.
	$$
	Then the dual unit ball of $\norm{\cdot}$ satisfies
	$$
		\cB_{\norm{\cdot}^\dagger}
		= \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}
	$$
	where $\sum$ denotes the Minkowski sum and $\cB_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

\begin{proof}
	Let us first prove the lemma for the case $n = 2$. Denote $f(x) = \alpha_1 \norm{x}_{(1)}$ and $g(x) = \alpha_2 \norm{x}_{(2)}$, such that $\norm{x} = f(x) + g(x)$. Recall two standard facts:
	\begin{enumerate}
		\item For any norm $\norm{\cdot}$ and $\lambda>0$,
		      $$
			      (\lambda \norm{\cdot})^*(y) =
			      \sup_{x}\bigl( \langle y,x \rangle - \lambda \norm{x}\bigr)
			      = \delta_{\lambda \cB_{\norm{\cdot}^\dagger}}(y)\,,
		      $$
		      i.e., the indicator function of the scaled dual ball.
		\item The Fenchel conjugate of a sum satisfies
		      $$
			      (f+g)^*(y) = \inf_{u+v=y} \bigl(f^*(u) + g^*(v)\bigr)\,.
		      $$
	\end{enumerate}
	Applying these to $f$ and $g$, we have
	$$
		f^*(u) = \delta_{\alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger}}(u)\,,
		\quad
		g^*(v) = \delta_{\alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}}(v)\,.
	$$
	Thus,
	$$
		\norm{\cdot}^*(y)
		= (f+g)^*(y)
		= \inf_{u+v=y}
		\bigl(
		\delta_{\alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger}}(u)
		+
		\delta_{\alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}}(v)
		\bigr)
		=
		\delta_{\alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger}+\alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}}(y).
	$$
	By definition, the conjugate of a norm is exactly the indicator of its dual unit ball: 
	$$
	\norm{\cdot}^*(y) = \delta_{\cB_{\norm{\cdot}^\dagger}}(y)\,.
	$$
	Therefore, $\cB_{\norm{\cdot}^\dagger} = \alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger} + \alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}$.

	Now we prove the general case by induction. The base case ($n=2$) is already proven. Suppose that the assumption of the lemma holds for $n = k$. Then, for $n = k + 1$, 
	$$
	\norm{x} = \sum_{i = 1}^k \alpha_i \norm{x}_{(i)} + \alpha_{k + 1}\norm{x}_{(k + 1)} = \norm{x}_{(1:k)} + \alpha_{k + 1}\norm{x}_{(k + 1)}\,.
	$$
	Applying the result for $n=2$ combined with the induction assumption, we obtain
	$$
	\cB_{\norm{\cdot}^\dagger} = \cB_{\norm{\cdot}_{(1:k)}^\dagger} + \alpha_{k + 1} \cB_{\norm{\cdot}_{(k + 1)}^\dagger} = \sum_{i = 1}^{k + 1} \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}\,,
	$$
	which proves the lemma.
\end{proof}

\begin{lemma}\label{lemma:lin_colm_alg}
	Let $\norm{\cdot}_{(1)}, \dots, \norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1, \dots, \alpha_n$ be non-negative reals. Consider Linear Minimization Oracles $\mathrm{LMO}_{1}, \dots, \mathrm{LMO}_{n}$, corresponding to these norms. Then, $\sum_{i = 1}^n \alpha_i \mathrm{LMO}_{i}$ is the LMO corresponding to the norm $\norm{\cdot}^\dagger$ dual to the $\sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

\begin{proof}
	From \cref{lemma:dual_to_conv_comb} we know the general form of the unit ball in the $\norm{\cdot}$ norm: $\cB_{\norm{\cdot}} = \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}$. Thus, the linear function minimization objective on a $\norm{\cdot}$ norm ball can be transformed as follows:
	$$
	\argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD> = \argmin_{\mD_1 \in \alpha_i \cB_i,\dots,\mD_n \in \alpha_n\cB_n}\<\mM, \sum_{i = 1}^n\mD_i> = \sum_{i = 1}^n \argmin_{D_i \in \cB_i}\<\mM, \mD_i>\,.
	$$
	This immediately implies $\sum_{i = 1}^n \alpha_i \mathrm{LMO}_{i} \in \argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD>$, which proves the claim of the lemma.
\end{proof}

\subsection{Frobeniusize Them: F-Muon and F-Neon}
Let us construct the concrete examples of algorithms given by linear combinations of LMO-algorithms. It follows from \cref{lemma:lin_colm_alg} that those algorithms can also be viewed as LMO-algorithms.

Consider a mixture of Fanions with another excellent LMO-algorithm, NSGD. We define $\normfstar{\cdot}$ as a convex combination of $\normn{\cdot}$ and $\normf{\cdot}$:

@@

\section{Computing the Updates}\label{sec:matrix-side}
To compute the algorithms' updates, we use thick-restarted Lanczos method for singular value problem (TRLan) on $\mM^{k\top}\mM^k$ or $\mM^k \mM^{k\top}$ matrices (the one with less size is picked), implemented in CuPy lSibrary \citep{cupy_svds_ref} and described in \citet{simonz1998thick}.
